Formulating the posterior for \tardis

### 2015-08-13 posterior/model.tex
[posterior] more structure thru subsections

\subsection{Extracting information from \tardis}

@@

\subsubsection{Running \tardis}
\todo{This is still rudimentary}
\begin{compactenum}[(I)]
  \item Adjust the temperature of an inner shell to approximately match
    the total luminosity inferred from the observations
  \item Choose a total number of real packets, and a number of virtual
    packets to send out on each scattering of a real packet.
  \item Run \tardis.
  \item Some of the real packets reenter the core and are thus
    unobservable. But $n^\real$ real and $n^\virt$ virtual packets
    leave the core.
  \item Each packet has a luminosity. The packets are put into the
    same $n_b$ bins as the observations such that
    \begin{align}
      \label{eq:bin-packets}
      n^\real = \sum_{i=1}^{n_b} n_i^\real
    \end{align}
  \item The sum of real luminosities in bin $i$ is $\lumi^\real$.
\end{compactenum}
I usually omit the superscript $\in \{\real, \virt\}$ on $n_i$,
$\lumi$, and dependent quantities but equations hold for both unless
noted otherwise.

\subsubsection{Statistical model}

@@

Let $N_i$ be the number of packets in we \emph{could} get in bin $i$
(it is not the number of packets we \emph{did} see, that's $n_i$),
then the law of total probability gives
\begin{equation}
  \label{eq:lum}
  P(\Lumi | \vecth) = \sum_{N_i=0}^{\infty} P(\Lumi | N_i, \vecth) P(N_i | \vecth)
\end{equation}
The first term in \refeq{lum} gives the luminosity in bin $i$ given
$N_i$ packets, so $\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$ is
the luminosity of packet $j$ in bin $i$. Obviously $P(\Lumi | 0,
\vecth) = \delta(\Lumi)$.

@@

With $P(r|\vecth)$ fixed, we can now look at sums luminosities. The
standard result is that if $r_1, r_2 \sim P(r|\vecth)$ then
\begin{equation}
  \label{eq:convolution}
  z_2 = r_1 + r_2 \sim \int \rmdx{r} P(r|\vecth) P(z_2 - r|\vecth) \,.
\end{equation}
Sums of $N_i$ terms are computed in identical fashion by convolution
of $P(r|\vecth)$ with the distribution of the sum with $N_i-1$
terms. For $N_i \gtrsim 15$, the central limit theorem applies and
\begin{equation}
  \label{eq:clt}
  P(\Lumi | N_i, \vecth) \approx \gaussian \left(\Lumi | N_i * E_{P(r|\vecth)}[r], \sqrt{N_i * V_{P(r|\vecth)}[r]} \right)
\end{equation}
Note that for the convolutions, we would have to use the approximate
\genbetapr{} distribution but for the mean and std. deviation in
\refeq{clt}, we should use the sample mean and std. deviation to
bypass the extra uncertainty induced by assuming $P(r | \vecth) =
\genbetapr(r | \vecnu(\vecth))$ and fixed $\vecnu$.

For $P(N_i | \vecth)$, the second term in \eqref{eq:lum}, we assume
$N_i$ follows a Poisson distribution whose rate parameter $\lambda_i =
\lambda_i(\vecth)$ has to be inferred from the simulation. We can use
the law of total probability
\begin{equation}
  P(N_i | \vecth) = \int \rmdx{\lambda_i} P(\lambda_i | \vecth) \poisson(N_i | \lambda_i, \vecth)
\end{equation}
\todo{postpone the integral over $\lambda$ then use result for
  compound Poisson distr. and finally integrate. Hopefully this can
  still all be done analytically} and Bayes' theorem
\begin{equation}
  P(\lambda_i | \vecth) \propto \poisson(n_i | \lambda_i) P_0(\lambda_i)
\end{equation}
If we choose Jeffrey's prior $P_0(\lambda_i) \propto
1/\sqrt{\lambda_i}$ then we get the analytical
result~\cite[Eq. 5]{Aggarwal:2011aa}
\begin{equation}
  P\given{N_i}{n_i} = \frac{n_i! [2 (N_i + n_i)]!}{4^{N_i} N_i! (2 n_i)! (N_i + n_i)!}
\end{equation}
where $n_i = n_i(\vecth)$ is the number of packets in bin $i$ actually
observed in the simulation at hand.  In essence, $P(N_i | \vecth)$ is
the posterior predictive distribution to have $N_i$ packets given that
the simulation spit out $n_i(\vecth)$ packages. So $n_i$ is known,
while $N_i$ is a random variable and we take into account its
uncertainty by summing over $N_i$ in \refeq{lum}. This is relevant for bins where $n_i$ is small. If we got
2 packets, it is not so unlikely we could have gotten one package. But
of course with only half the packages, the expectation for the sum of
luminosities changes significantly.

>>>>

@@

\subsection{Running \tardis}
\todo{This is still rudimentary}
\begin{compactenum}[(I)]
  \item Adjust the temperature of an inner shell to approximately match
    the total luminosity inferred from the observations
  \item Choose a total number of real packets, and a number of virtual
    packets to send out on each scattering of a real packet.
  \item Run \tardis.
  \item Some of the real packets reenter the core and are thus
    unobservable. But $n^\real$ real and $n^\virt$ virtual packets
    leave the core.
  \item Each packet has a luminosity. The packets are put into the
    same $n_b$ bins as the observations such that
    \begin{align}
      \label{eq:bin-packets}
      n^\real = \sum_{i=1}^{n_b} n_i^\real
    \end{align}
  \item The sum of real luminosities in bin $i$ is $\lumi^\real$.
\end{compactenum}
I usually omit the superscript $\in \{\real, \virt\}$ on $n_i$,
$\lumi$, and dependent quantities but equations hold for both unless
noted otherwise.

\subsection{Statistical model}

@@

Let $N_i$ be the number of packets in we \emph{could} get in bin $i$
(it is not the number of packets we \emph{did} see, that's $n_i$),
then the law of total probability gives
\begin{equation}
  \label{eq:lum}
  P(\Lumi | \vecth) = \sum_{N_i=0}^{\infty} P(\Lumi | N_i, \vecth) P(N_i | \vecth)
\end{equation}

\subsubsection{Luminosity distributions}

The first term in \refeq{lum} gives the luminosity in bin $i$ given
$N_i$ packets, so $\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$
is the luminosity of packet $j$ in bin $i$. Obviously $P(\Lumi | 0,
\vecth) = \delta(\Lumi)$. The crucial term is
$P\given{L_i}{N_i=1,\vecth}$: it is the distribution of luminosity of
a single packet in bin $i$ and determines all other terms with
$N_i>1$.

@@

With $P(r|\vecth)$ fixed, we can now look at sums of luminosities. The
standard result is that if $r_1, r_2 \sim P(r|\vecth)$ then
\begin{equation}
  \label{eq:convolution}
  z_2 = r_1 + r_2 \sim \int \rmdx{r} P(r|\vecth) P(z_2 - r|\vecth) \,.
\end{equation}
Sums of $N_i$ terms are computed in identical fashion by convolution
of $P(r|\vecth)$ with the distribution of the sum with $N_i-1$
terms. For $N_i \gtrsim 15$, the central limit theorem applies and
\begin{equation}
  \label{eq:clt}
  P(\Lumi | N_i, \vecth) \approx \gaussian \left(\Lumi | N_i * E_{P(r|\vecth)}[r], \sqrt{N_i * V_{P(r|\vecth)}[r]} \right)
\end{equation}
Note that for the convolutions, we would have to use the approximate
\genbetapr{} distribution but for the mean and std. deviation in
\refeq{clt}, we should use the sample mean and std. deviation to
bypass the extra uncertainty induced by assuming $P(r | \vecth) =
\genbetapr(r | \vecnu(\vecth))$ and fixed $\vecnu$.

\subsubsection{Poisson noise}

Let us now consider $P(N_i | \vecth)$, the second term in
\eqref{eq:lum}. As stated before, we assume the number of packets in
the given simulation follows a Poisson distribution whose rate
parameter $\lambda_i = \lambda_i(\vecth)$ has to be inferred from the
observed $n_i$. We can use the law of total probability
\begin{equation}
  P(N_i | \vecth) = \int \rmdx{\lambda_i} P(\lambda_i | \vecth) \poisson(N_i | \lambda_i, \vecth)
\end{equation}
\todo{postpone the integral over $\lambda$ then use result for
  compound Poisson distr. and finally integrate. Hopefully this can
  still all be done analytically} and Bayes' theorem
\begin{equation}
  P(\lambda_i | \vecth) \propto \poisson(n_i | \lambda_i) P_0(\lambda_i)
\end{equation}
If we choose Jeffrey's prior $P_0(\lambda_i) \propto
1/\sqrt{\lambda_i}$ then we get the analytical
result~\cite[Eq. 5]{Aggarwal:2011aa}
\begin{equation}
  P\given{N_i}{n_i} = \frac{n_i! [2 (N_i + n_i)]!}{4^{N_i} N_i! (2 n_i)! (N_i + n_i)!}
\end{equation}
where $n_i = n_i(\vecth)$ is the number of packets in bin $i$ actually
observed in the simulation at hand.  In essence, $P(N_i | \vecth)$ is
the posterior predictive distribution to have $N_i$ packets given that
the simulation spit out $n_i(\vecth)$ packages. So $n_i$ is known,
while $N_i$ is a random variable and we take into account its
uncertainty by summing over $N_i$ in \refeq{lum}. This is relevant for bins where $n_i$ is small. If we got
2 packets, it is not so unlikely we could have gotten one package. But
of course with only half the packages, the expectation for the sum of
luminosities changes significantly.
